---
title: "census_download_scripts"
format: html
editor: visual
execute: 
  warning: false
  message: false
---

# Download data from Census Bureau

## Setup

```{r}

library(tidycensus) 
library(tigris) 
library(sf) 
library(dplyr) 
library(ggplot2) 
library(readr) 
library(tidyr) 
options(tigris_use_cache = TRUE)

```

## 1 API KEY

> Run this only once IN THE CONSOLE DO NOT ADD API KEY TO THE SCRIPT

`census_api_key("YOUR API KEY HERE", install = TRUE, overwrite = TRUE)`

> ATTENTION! API KEYS expires quickly. Request a new one whenever needed.

## 2 Explore variables

Download a complete list of codes and descriptions available for the 2024 ACS 5-year ("acs5") and stores it in a table called vars. It is like a data dictionary

The Census Bureau has thousands of variables in the ACS, and each one is identified by a cryptic code like B19013_001 or B17001_002 — nobody memorizes these. load_variables() downloads the full catalog of available variable codes and their human-readable descriptions (like a dictionary or index) and stores it in vars.

This step doesn't download any actual data about counties, states, income, or poverty numbers. It's purely a reference table you browse through (you filtered it with grepl("\^B19", name)) to figure out which exact codes correspond to the concepts you want to study — in your case, median household income and poverty. You use this step once, just to look things up, then you already know the codes you need (B19013_001, B17001_002, B17001_001), which get hardcoded into my_vars in the next part of the script.

```{r}

vars <- load_variables(2023, "acs5")

```

Keeps only the rows where the variable name starts with "B19".

- `grepl` is a text/pattern search (regex), and `\^B19` means "starts with B19".

Variables starting with B19 are all income-related, so this filters the giant dictionary down to just the income variables.

- `slice_head` can also capture the top n rows of each group with `group_by()`

```{r}

vars <- dplyr::filter(vars, grepl("^B19", name)) %>% 
  dplyr::slice_head(n = 10) 
print(vars)

```

## 3 Parameters

Now that you know exactly which variable codes you want, get_acs() is the function that actually contacts the Census API and pulls the real numbers for those specific codes, for the specific geography you chose (here, counties in Missouri). This is the step that returns actual estimates and margins of error for real places, plus (because geometry = TRUE) the map boundaries for each county.

But before, let's set up the paramers for our search.

```{r}

survey <- "acs5" 
year_acs <- 2024 
state_abbr <- "MO" 
geo_level <- "county" 
my_vars <- c( 
  income = "B19013_001", # median household income 
  pov = "B17001_002", # people below poverty 
  povden = "B17001_001" # poverty universe = denominator 
  )

```

## 4 Download the data

Let's now download the data

```{r}

acs_wide <- get_acs( 
  geography = geo_level, 
  variables = my_vars, 
  state = state_abbr, 
  year = year_acs, 
  survey = survey, 
  geometry = TRUE, 
  # 'wide' for variables in columns 'long' is default for variables in rows 
  output = "wide" 
  )
                     
```

In R, an object can be more than one thing at the same time:

- `data.frame` means it's a table

- `sf` means it's also a spatial object (simple features) because it has a special column storing the geometry

- That extra layer is what lets you use `geom_sf()` in `ggplot()` to draw maps

```{r}

class(acs_wide)

```

## 5 Derive a rate, and carry its margin of error through

The chunk below calculates the **poverty rate** for each county: it takes the estimated number of people below the poverty line (`povE`, from `B17001_002`) and divides it by the total population in the "poverty universe" for that county (`povdenE`, from `B17001_001` — the denominator, i.e., the population this poverty measure applies to). The result is a proportion (for example, 0.15 = 15% of people in that county are below the poverty line).

When you simply divide two numbers that each already have their own margins of error (`povM` and `povdenM`), you **can't** just divide the margins of error too (`povM / povdenM` would be wrong). Because `povE` (people in poverty) is a subset of `povdenE` (everyone), the two estimates are statistically correlated, and there's a specific Census Bureau formula for correctly calculating the margin of error of a **proportion** derived from two other estimates.

`moe_prop()` is a function from the `tidycensus` package itself that already implements that official Census formula for you. It takes: - `povE` — numerator estimate (people in poverty) - `povdenE` — denominator estimate (total population) - `povM` — numerator's MOE - `povdenM` — denominator's MOE

And returns the correct MOE for the `pov_rate` you just calculated.

In practice: without this line, you'd have each county's poverty rate but no idea how reliable that number is. With `pov_moe`, you can say, for example, "this county's poverty rate is 15% ± 3 percentage points" — which is essential for the part of the assignment that asks you to compare areas and check whether their confidence intervals overlap (the "Uncertainty check" deliverable).

```{r}

acs_wide <- acs_wide %>% 
  mutate(
    pov_rate = povE / povdenE, 
    pov_moe = moe_prop(povE, povdenE, povM, povdenM)
    )

```

## 6 Map (edit titles/theme)

```{r}

ggplot(acs_wide) + 
geom_sf(aes(fill = incomeE), color = NA) + 
  scale_fill_viridis_c(name = "Median HH Income") + 
  labs(title = paste0("ACS ", 
                      year_acs, 
                      " 5-year: Median Income — ", 
                      state_abbr, 
                      " (", 
                      geo_level, 
                      ")"
                      ), 
       caption = "Source: U.S. Census Bureau via tidycensus") + 
  theme_minimal() + 
  theme(#text = element_text(family = "Palatino"),
        plot.title = element_text(size = 14, face = "bold"), 
        plot.caption = element_text(size = 8), 
        legend.position.inside = c(0.2, 0.15)
  )

```

## 7 Table (top/bottom by poverty rate, with MOE)

```{r}

top10 <- acs_wide %>% 
  st_drop_geometry() %>% 
  arrange(desc(pov_rate)) %>% 
  select(NAME, pov_rate, pov_moe, povE, povM) %>% 
  slice_head(n = 10) 
bottom10 <- acs_wide %>% 
  st_drop_geometry() %>% 
  arrange(pov_rate) %>% 
  select(NAME, pov_rate, pov_moe, povE, povM)  %>% 
  slice_head(n = 10) 
top10 
bottom10

```

## 8 Save outputs

```{r}

write_csv(st_drop_geometry(acs_wide), paste0("acs_", state_abbr, "_", year_acs, ".csv"))

```
