---
title: "assign02_Data_Visualization"
format: html
editor: visual
execute:
  warning: false
  message: false
---

## Exercise 1 - Run Murrell's R Graphics code

> Unable to complete exercise 1. The file `murrell01.R` was not uploaded to the Teams Folder until the day I did the exercise (Sep 19/2026)

## Exercise 2

Using the dataset `mtcars`

```{r}

df <- mtcars
```

### a) Set up the canvas

- `par()` sets global graphical parameters (screen layout, margins, default colors) that apply to all plots drawn afterward, until you change them again.

```{r}

plot.new()                             # opens a blank canvas
#plot.window(xlim = c(0, 10), ylim = c(0, 10))
box()

```

- `box()` draws a frame around the plotting area

```{r}

par(mfrow = c(1, 2))   # splits the "screen" into 1 row, 2 columns
plot.new()
box()
plot.new()# opens a blank canvas
box()

```

- `axis()` manually draws the axes (tick marks and labels) on a specific side of the plot

```{r}

par(mfrow = c(1, 2))   # splits the "screen" into 1 row, 2 columns
plot.new()
plot.window(xlim = c(0, 10), ylim = c(0, 10))
box()
axis(1)
axis(2)
plot.new()
plot.window(xlim = c(0, 10), ylim = c(0, 10))
box()
axis(1)
axis(2)

```

### b) Draw whole plots

`hist()` draws a histogram, showing the frequency distribution of a numeric variable.

```{r}

hist(mtcars$wt)

```

`boxplot()` draws a box plot, showing the summary (median, quartiles, outliers) of a variable, and can compare groups.

```{r}

boxplot(mtcars$mpg)

```

`pie()` draws a pie chart, showing the proportion of each category within a total.

```{r}

pie(table(mtcars$cyl))

```

`persp()` draws a 3D surface, showing how a matrix of values (height z over an x-y grid) is represented in space.

```{r}

persp(volcano)

```

### c) Add to an existing plot

`points()` adds extra points to an existing plot, without erasing what was already drawn.

```{r}

plot(mtcars$wt, mtcars$mpg)
points(mtcars$wt[mtcars$cyl == 8], mtcars$mpg[mtcars$cyl == 8],
       col = "red", pch = 19)   # highlights the 8-cylinder cars in red

```

`lines()` adds a line (connecting points) to an existing plot.

```{r}

plot(mtcars$wt, mtcars$mpg)
lines(lowess(mtcars$wt, mtcars$mpg), col = "blue")   # adds a trend line

```

### d) Annotate

`text()` adds free-standing text inside the plotting area, at specific coordinates.

```{r}

plot(mtcars$wt, mtcars$mpg)
text(mtcars$wt, mtcars$mpg, labels = rownames(mtcars), pos = 3, cex = 0.6)

```

`mtext()` adds text in the plot margin (outside the plotting area), controlling which side/line of the margin receives the text.

```{r}

plot(mtcars$wt, mtcars$mpg)
text(mtcars$wt, mtcars$mpg, labels = rownames(mtcars), pos = 3, cex = 0.6)
mtext("Weight vs. Fuel Consumption — mtcars", side = 3, line = 1)

```

`legend()` draws a legend explaining what each color/symbol in the plot represents.

```{r}

plot(mtcars$wt, mtcars$mpg, col = as.factor(mtcars$cyl), pch = 19)
text(mtcars$wt, mtcars$mpg, labels = rownames(mtcars), pos = 3, cex = 0.6)
mtext("Weight vs. Fuel Consumption — mtcars", side = 3, line = 1)
legend("topright", legend = c("4 cyl", "6 cyl", "8 cyl"),
       col = 1:3, pch = 19)

```

`names()` is not a graphical function — it reads or sets names (of vector elements, data frame columns, etc.), making it part of data preparation rather than plotting.

```{r}

names(mtcars)              # shows the column names
names(mtcars)[1] <- "miles_per_gallon"   # renames the first column
```

## Exercise 3

```{r}
library(readxl)
dataset <- read_xlsx('Happy-Planet-Index-2006-2025-public-data-set.xlsx',
                     sheet=2, skip = 8)

plot(dataset$`GDP per capita`, dataset$`Life satisfaction`, col = 'blue',
     main = "Correlation between GDP per capita and Life Satisfaction",
     xlab = "GDP per capita",
     ylab = "Life Satisfaction")
mtext("Source: https://happyplanetindex.org/countries/",
      side = 1, line = 4, cex = 0.7, adj = 1)

```
