---
format: html
editor: visual
execute:
  warning: false
  message: false
---
# Assignment 3 - EPPS 6355 - Data Visualization

## 1. Document a Chart

```{r}

library(ggplot2)

# Creating the data frame with the data: proportion of prefered pie flavors)
sales <- data.frame(
  flavor = c("Blueberry", "Cherry", "Apple", "Boston Cream", "Other", "Vanilla"),
  prop  = c(0.12, 0.3, 0.26, 0.16, 0.04, 0.12)
)

# Keeping the original order of the flavores (to keep ggplot from sorting 
# alphabetically)
sales$sabor <- factor(sales$flavor, levels = sales$flavor)

# Plotting
# Ggplot2 has no pie chart. We need to create a col chart and roll it in
ggplot(sales, aes(x = 1, y = prop, fill = flavor)) +
# Plotting a single stacked bar
  geom_col(width = 1, color = "black",
           position = position_stack(reverse = TRUE)) +
# Placing each flavor's name on the outside and middle of each slice
  geom_text(aes(x = 1.7, label = sabor),
            position = position_stack(vjust = 0.5, reverse = TRUE), size = 3) +
  # Wrapping the bar in a circle, from 3 o'clock and spinning anti-clock
  coord_polar(theta = "y", start = pi / 2, direction = -1) +
  # Using 6 shades of gray from 0.3 to 1 where 1 is totally white
  scale_fill_manual(values = gray(seq(0.3, 1.0, length = 6))) +
  # Removes the axes, the background and legends
  theme_void() +
  theme(legend.position = "none")

```

## 2. Anscombe's quartet
Anscombe's quartet shows that different data sets can lead to the same modeled outcomes even though the data sets behave completely differently. From the previous exercise on Anscombe's quartet, we have the following plots (I added the slope, y-intercept, Rsquare and Residual Standard Error for better comparison).

```{r}
## Anscombe (1973) Quartlet
data(anscombe)  # Load Anscombe's data
summary(anscombe)

## Regression models
lm1 <- lm(y1 ~ x1, data=anscombe)
summary(lm1)

lm2 <- lm(y2 ~ x2, data=anscombe)
summary(lm2)

lm3 <- lm(y3 ~ x3, data=anscombe)
summary(lm3)

lm4 <- lm(y4 ~ x4, data=anscombe)
summary(lm4)

## Plotting the regressions
ff <- y ~ x
mods <- setNames(as.list(1:4), paste0("lm", 1:4))

# Plot using for loop
for(i in 1:4) {
  ff[2:3] <- lapply(paste0(c("y","x"), i), as.name)
  ## or   ff[[2]] <- as.name(paste0("y", i))
  ##      ff[[3]] <- as.name(paste0("x", i))
  mods[[i]] <- lmi <- lm(ff, data = anscombe)
  print(anova(lmi))
}

sapply(mods, coef)  # Note the use of this function
lapply(mods, function(fm) coef(summary(fm)))

# Extracting coefficients, R-squared, and Residual Standard Error (RSE)
intercepts <- sapply(mods, function(fm) coef(fm)[1])
slopes     <- sapply(mods, function(fm) coef(fm)[2])
r2_vals    <- sapply(mods, function(fm) summary(fm)$r.squared) # R2
rse_vals   <- sapply(mods, function(fm) summary(fm)$sigma)     # Residual Standard Error

# Preparing for the plots
op <- par(mfrow = c(2, 2), mar = 0.1+c(4,4,1,1), oma =  c(0, 0, 2, 0))

# Plot charts using for loop with parameters printed on each plot
for(i in 1:4) {
  ff[2:3] <- lapply(paste0(c("y","x"), i), as.name)
  plot(ff, data = anscombe, col = "red", pch = 21, bg = "tomato", cex = 1.2,
       xlim = c(3, 19), ylim = c(3, 13))
  abline(mods[[i]], col = "blue")
  
  # Print Intercept, Slope, R2, and RSE inside each plot window
  text(x = 3.5, y = 12.5, labels = paste0("Intercept = ", round(intercepts[i], 3)), pos = 4, cex = 0.85)
  text(x = 3.5, y = 11.7, labels = paste0("Slope = ", round(slopes[i], 3)), pos = 4, cex = 0.85)
  text(x = 3.5, y = 10.9, labels = paste0("R2 = ", round(r2_vals[i], 3)), pos = 4, cex = 0.85)
  text(x = 3.5, y = 10.1, labels = paste0("RSE = ", round(rse_vals[i], 3)), pos = 4, cex = 0.85)
}
mtext("Anscombe's 4 Regression data sets", outer = TRUE, cex = 1.5)


```


>We can see that the values for the slope, the intercept, the R2 and the RSE are the same for all four models. That reveals that without visualizing the data, an analyst could make serious modeling mistakes. Plotting the data, however reveals that each data set behaves very differently from the others. Data set 2 and data set 4, for example, are clearly nonlinear, which is prohibitive for regressions.

Another way to see the regressions is using confidence intervals. Once again, a different visualization proves that the data sets are different, regardless of the equivalent metrics:

```{r}

library(ggplot2)
library(patchwork)

data(anscombe)

plot_anscombe <- function(dataset_num) {
  x_col <- paste0("x", dataset_num)
  y_col <- paste0("y", dataset_num)
  formula <- as.formula(paste(y_col, "~", x_col))
  mod <- lm(formula, data = anscombe)
  intercept <- round(coef(mod)[1], 3)
  slope     <- round(coef(mod)[2], 3)
  r2        <- round(summary(mod)$r.squared, 3)
  rse       <- round(summary(mod)$sigma, 3)
  stats_label <- paste0(
    "Intercept = ", intercept, "\n",
    "Slope = ", slope, "\n",
    "R² = ", r2, "\n",
    "RSE = ", rse
  )
  
  # Building confidence bands
  ggplot(anscombe, aes_string(x = x_col, y = y_col)) +
    geom_smooth(method = "lm", formula = y ~ x, se = TRUE, 
                color = "blue", fill = "lightblue", alpha = 0.4) +
    geom_point(fill = "tomato", color = "red", shape = 21, size = 3) +
    annotate("text", x = 3.5, y = 11.5, label = stats_label, 
             hjust = 0, size = 3.2, family = "sans") +
    scale_x_continuous(limits = c(3, 19)) +
    scale_y_continuous(limits = c(3, 13)) +
    labs(title = paste("Dataset", dataset_num), x = paste0("x", dataset_num), y = paste0("y", dataset_num)) +
    theme_bw()
}

p1 <- plot_anscombe(1)
p2 <- plot_anscombe(2)
p3 <- plot_anscombe(3)
p4 <- plot_anscombe(4)

(p1 | p2) / (p3 | p4) + 
  plot_annotation(
    title = "Anscombe's 4 Regression Data Sets (with 95% CI Bands)",
    theme = theme(plot.title = element_text(size = 16, face = "bold", hjust = 0.5))
  )

```



## 3. Finetune without leaving base R

Importing data and running regression with data set 1 from 
```{r}

## Stacked Barplot (VADeaths) - Custom Styled

data(VADeaths)

# Set serif font family globally
par(family = "serif", mar = c(3, 4, 3, 2))

# Non-default color palette for the 5 age categories
custom_cols <- c("#2B5C8F", "#4682B4", "#68BB68", "#E69F00", "#D55E00")

# Draw stacked barplot
midpts <- barplot(VADeaths, 
                  col = custom_cols, 
                  names = rep("", 4),
                  main = "Death Rates in Virginia (1940)",
                  ylab = "Deaths per 1000",
                  col.main = "darkblue",
                  col.lab = "gray10",
                  col.axis = "gray20",
                  ylim = c(0, 225))

# Add x-axis labels
mtext(sub(" ", "\n", colnames(VADeaths)),
      at = midpts, side = 1, line = 1, cex = 0.8)

# Add numeric text inside the stacked segments
text(rep(midpts, each = 5), 
     apply(VADeaths, 2, cumsum) - VADeaths / 2,
     VADeaths, 
     col = "white", 
     font = 2,
     cex = 0.8)

# Custom plotting character (pch = 23: filled diamond) placed at total bar heights
total_heights <- colSums(VADeaths)
points(midpts, total_heights + 6, 
       pch = 23, bg = "#D55E00", col = "black", cex = 1.4, lwd = 1.2)

```

## 4. Now in ggplot2

```{r}

library(ggplot2)
library(tidyr)
library(dplyr)

# 1. Preparação dos dados em formato longo (tidy format)
data(VADeaths)

df_vadeaths <- as.data.frame(VADeaths) %>%
  mutate(Age = rownames(VADeaths)) %>%
  pivot_longer(cols = -Age, names_to = "Group", values_to = "Deaths") %>%
  mutate(
    Age = factor(Age, levels = rev(rownames(VADeaths))), # Mantém a ordem do empilhamento
    Group = sub(" ", "\n", Group)
  )

# Totais por grupo para posicionar o caracter de plotagem no topo de cada barra
df_totals <- df_vadeaths %>%
  group_by(Group) %>%
  summarise(Total = sum(Deaths))

# Paleta de cores personalizada (não padrão)
custom_cols <- c("#D55E00", "#E69F00", "#68BB68", "#4682B4", "#2B5C8F")

# 2. Plotagem com ggplot2
ggplot(df_vadeaths, aes(x = Group, y = Deaths, fill = Age)) +
  # Barras empilhadas com borda fina
  geom_bar(stat = "identity", position = "stack", color = "white", width = 0.6) +
  # Texto numérico dentro dos segmentos das barras
  geom_text(aes(label = Deaths), 
            position = position_stack(vjust = 0.5), 
            color = "white", fontface = "bold", size = 3.5, family = "serif") +
  # c. Own plotting character (Diamantes/pch=23 no topo das barras)
  geom_point(data = df_totals, aes(x = Group, y = Total + 6, fill = NULL), 
             shape = 23, fill = "#D55E00", color = "black", size = 3.5, stroke = 1) +
  # b. Non-default colours
  scale_fill_manual(values = custom_cols) +
  scale_y_continuous(limits = c(0, 225)) +
  labs(
    title = "Death Rates in Virginia (1940)",
    x = NULL,
    y = "Deaths per 1000",
    fill = "Age Group"
  ) +
  theme_minimal() +
  # a. Use a serif font globally
  theme(
    text = element_text(family = "serif"),
    plot.title = element_text(face = "bold", size = 14, color = "darkblue", hjust = 0.5),
    axis.title.y = element_text(size = 11, color = "gray10"),
    axis.text = element_text(size = 10, color = "gray20"),
    legend.position = "right"
  )

```

## 5. Pre‑hackathon — team task

#### Setting up libraries and downloading Daily COVID counts, Our World in Data. 

```{r}
library(ggplot2)
library(tidyverse)

# Daily COVID counts, Our World in Data.
url <- paste0("https://raw.githubusercontent.com/owid/covid-19-data/",
              "master/public/data/owid-covid-data.csv")
download.file(url, "owid-covid-data.csv")
owidall <- read.csv("owid-covid-data.csv")

# Drop OWID aggregate rows (continents, income groups, World)
owidall <- owidall[!grepl("^OWID", owidall$iso_code), ]

# Subset by continent
owideu <- subset(owidall, continent == "Europe")
owideu$date <- as.Date(owideu$date)

# Removing Russia since it seems to be absent in the graph template
df <- dplyr::filter(owideu,
                    !is.na(new_deaths),
                    location != "Russia")
```

#### Calculating the quartiles and IQR for outliers

```{r}

q3 <- quantile(df$new_deaths, 0.75, na.rm=T)
iqr <- IQR(df$new_deaths)
sup <- q3 + 1.5*iqr
df$outliers <- ifelse(df$new_deaths > sup, df$location, NA)

```

#### Plotting

```{r}

ggplot(df)+
  aes(x=date, y=new_deaths)+
  geom_point(col='#E7298A')+
  #geom_text(aes(label=outliers), na.rm=T)+
  scale_x_date(
    date_breaks = "2 months",
    date_labels = "%Y-%m"
    )+
  labs(title = "Daily COVID deaths in Europe",
       y = "COVID Deaths in Europe (Daily)",
       x = "Date")+
  #scale_y_continuous(limits=c(0,6600))+
  theme_bw()+
  theme(
    axis.text.x = element_text(angle=90, vjust=0.5, hjust=1),
    panel.grid = element_blank()
  )

```