---
format: html
editor: visual
execute: 
  warning: false
  message: false
---

# Assignment 02 - EPPS 6302

This assignment addresses term searches on Google Trends

> The following notes reproduce the scripts used for each search, an overview of the scraped data and the respective plots.

------------------------------------------------------------------------

#### Installing needed packages

```{r}
library(gtrendsR)
library(readr)
library(DT)
```

------------------------------------------------------------------------

## Initial search

##### Query record

- Terms: Trump, Harris, election

- Countries: United States

- Group: Web

- Time window: Last 5 years

- Category: All categories (news, youtube, images etc)

```{r}
TrumpHarrisElection = gtrends(c("Trump","Harris","election"), 
                              onlyInterest = TRUE, 
                              geo = "US", 
                              gprop = "web", 
                              time = "today+5-y", # last five years
                              category = 0 
) 
df=TrumpHarrisElection$interest_over_time
plot(TrumpHarrisElection)
DT::datatable(TrumpHarrisElection$interest_over_time)
```

------------------------------------------------------------------------

## Updated Search

##### Query record

- Terms: immigration

- Countries: all

- Group: alll

- Time window: alll

- Category: all

```{r}
query2 <- gtrends("immigration", time = "all")
plot(query2)
DT::datatable(query2$interest_over_time)
```

------------------------------------------------------------------------

## Updated Search

##### Query record

- Terms: immigration

- Countries: United States, Great Britain, Taiwan

- Group: alll

- Time window: alll

- Category: all

```{r}
query3 <- gtrends("immigration", geo = c("US", "GB", "TW"), time = "all")
plot(query3)
DT::datatable(query3$interest_over_time)
```
